---
title: "Coursera DS 5 / Course Project 1"
author: "Jim Starks"
date: "August 4, 2018"
output: html_document
---


## Load libraries and Set Options
```{r}
library(ggplot2)
library(scales)
library(data.table)
theme_set(theme_minimal())
```

##Code for reading in the dataset and processing the data

```{r}
activity<-read.csv("activity.csv",sep=",",header=TRUE)
activity$dayofweek<-weekdays(as.Date(activity$date))
activity$weekend[activity$dayofweek %in% c('Saturday','Sunday')]<-"Weekend"
activity$weekend[is.na(activity$weekend)]<-"Weekday"

```


##Histogram of the total number of steps taken each day

```{r}
stepsPerDay<-tapply(activity$steps, activity$date, FUN=sum,na.rm=TRUE)
hist(stepsPerDay, main="Histogram of Average Steps per Day by Interval (missing values excluded)")
```

##Mean and median number of steps taken each day
```{r}
mean(stepsPerDay)
median(stepsPerDay)
```


##Create dataset to aggregate across all days by time interval
```{r}
avgStepsInterval<-aggregate(activity$steps, list(activity$interval), FUN=mean,na.rm=TRUE)
colnames(avgStepsInterval)<-c("interval","steps")

hour<-sprintf("%02d",floor(avgStepsInterval[,1]/100))
minute<-sprintf("%02d",avgStepsInterval[,1]-floor(avgStepsInterval[,1]/100)*100)
avgStepsInterval$hmTime<-strptime(paste(hour, minute, sep=":"),"%H:%M")

avgStepsInterval$hmTime<-as.POSIXct(avgStepsInterval$hmTime)
```

##Time series plot of the average number of steps taken

```{r}
ggplot(data = avgStepsInterval, aes(x = hmTime, y=steps))+
  geom_line(color = "#00AFBB", size = 2)+
  scale_x_datetime("",labels = date_format("%H:%M"))
```  

##The 5-minute interval that, on average, contains the maximum number of steps

```{r}
avgStepsInterval[which(avgStepsInterval[,2]==max(avgStepsInterval[,2])),]
```
##Code to describe and show a strategy for imputing missing data

###Find the total number of rows with NA's
```{r}
sum(is.na(activity$steps))
```

###impute missing values by weekday/weekend and time interval
```{r}
avg_steps_bytime <-aggregate(activity$steps, by=list(activity$weekend,activity$interval), FUN=mean, na.rm=TRUE)
colnames(avg_steps_bytime)<-c("weekend", "interval", "avgsteps")
```

###Check that there are no missing values in steps to be imputed

```{r}
sum(is.na(avg_steps_bytime$avgsteps))
```

###Create dataset with imputed missing values

```{r}
activity_nomiss<-activity
activity_nomiss<-data.table(activity_nomiss)
avg_steps_bytime<-data.table(avg_steps_bytime)
keys <- c("weekend","interval")
setkeyv(activity_nomiss, keys)
setkeyv(avg_steps_bytime, keys)

activity_nomiss<-merge(activity_nomiss, avg_steps_bytime)
activity_nomiss$avgsteps<-NULL
```

##Histogram of the total number of steps taken each day after missing values are imputed

```{r}
stepsPerDay_nomiss<-tapply(activity_nomiss$steps, activity$date, FUN=sum,na.rm=TRUE)
hist(stepsPerDay_nomiss, main="Histogram of Average Steps per Day by Interval (missing values imputed)")
mean(stepsPerDay_nomiss)
median(stepsPerDay_nomiss)
```

##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
avgStepsInterval2<-aggregate(activity$steps, list(activity$interval, activity$weekend), FUN=mean,na.rm=TRUE)
colnames(avgStepsInterval2)<-c("interval","weekend", "steps")

hour<-sprintf("%02d",floor(avgStepsInterval2[,1]/100))
minute<-sprintf("%02d",avgStepsInterval2[,1]-floor(avgStepsInterval2[,1]/100)*100)
avgStepsInterval2$hmTime<-strptime(paste(hour, minute, sep=":"),"%H:%M")

avgStepsInterval2$hmTime<-as.POSIXct(avgStepsInterval2$hmTime)

ggplot(data = avgStepsInterval2, aes(x = hmTime, y=steps))+
  geom_line(color = "#00AFBB", size = 2)+
  scale_x_datetime("",labels = date_format("%H:%M"))+
  facet_wrap(~weekend, ncol=1)
```  

